---
title: "Data Containers and Data Visualization"
author: "Niels Richard Hansen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Data Containers and Data Visualization}
  %\VignetteEncoding{UTF-8}
---

```{r init, echo=FALSE, message=FALSE, warning=FALSE}
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(dpi = 144, fig.width= 6, fig.height = 4, out.width = "90%", fig.align = "center")
```

This vignette demonstrates how to use the classes `ContinuousProcess`,
`MarkedPointProcess` and `JumpProcess` as data containers and how 
to visualize the data they contain.

First we load the example data. The data sets are `contExam` of class
`ContinuousProcess`, `pointExam` of class `MarkedPointProcess` and
`jumpExam` of class `JumpProcess`.

```{r }
library(ppstat)
data(example)
```

Showing data structure. 

```{r }
contExam
```

Summarizing the data.

```{r }
summary(contExam)
```

Plotting data.

```{r }
plot(contExam)
```

Subsetting data.

```{r }
subset(contExam, id %in% c("C","D","F"))
subset(contExam, foo > 0)
subset(contExam, gender == "F")
```

Plotting a subset.

```{r }
plot(subset(contExam, subset = (time > 4 & gender == "F"), select = foo))
```

Modifying plot using ggplot2.

```{r }
plot(subset(contExam, id == "A")) + geom_point()
plot(contExam) + facet_null()
```

More sophisticated plots can be produced, e.g. different uses of
faceting and grouping.

```{r }
plot(contExam) + facet_grid(variable ~ ., scale = "free_y")
```

To do this depending on unit-specific variables we must specify an
additional argument to the plot such that these variables are
included in the plot data.

```{r }
plot(contExam, allUnitData = TRUE) +
  facet_grid(variable ~ gender, scale = "free_y")
```

Subsetting to non-contiguous subsets of the time axis is
possible, but `plot` assumes a contiguous x-axis and interpolates
the graph.

```{r }
plot(subset(contExam, id == "A" & (time < 2 | time > 4)))
```

Factor columns are plotting using a different aesthetic mapping.

```{r }
plot(factExam)
```

The default is to drop one level (the first), but this can be
changed or bypassed by setting the `dropLevels argument.

```{r }
plot(factExam, dropLevels = 2)
plot(factExam, dropLevels = NULL)
```

Summarizing a MarkedPointProcess object gives additional
informations on the number of points for each unit and each mark. 

```{r }
summary(pointExam)
```

A combined plot for unit "A". 

```{r }
plot(subset(pointExam, id =="A"))
```

Plotting the point process data only.

```{r }
plot(pointExam[, -(1:3)])
```

Or a slightly different version.

```{r }
plot(pointExam[, -(1:3)], y = "id") + facet_null()
```

Plotting the continuous process data only.

```{r }
plot(pointExam[, -c(4,5)])
```

or 

```{r }
plot(as(pointExam, "ContinuousProcess"))
```

Some modifications of the plot.

```{r }
plot(pointExam, allUnitData = TRUE) + facet_grid(.~gender)
plot(pointExam, y = "id", allUnitData = TRUE) + facet_grid(variable~gender, scale = "free_y")
plot(pointExam, y = "foo") 
plot(pointExam, y = "@top", allUnitData = TRUE) + facet_grid(.~gender)
plot(pointExam, y = "@bottom", allUnitData = TRUE) + facet_grid(gender~., scales = "free_y")
plot(pointExam, y = "id") + facet_null()
```

A `JumpProcess` data set has points with the same marks as the variable names
in the `ContinuousProcess` part of the data set. In addition, to each point
there is a size of the jump associated.

```{r }
jumpExam
plot(jumpExam)
```

We can again use ggplot2 for modifications

```{r }
plot(jumpExam) + facet_grid(variable ~ .) + aes(color = id)
```

Removing objects.

```{r }
rm(contExam)
rm(pointExam)
rm(jumpExam)
```

