---
title: "Toy Models"
author: "Niels Richard Hansen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Toy Models}
  %\VignetteEncoding{UTF-8}
---

```{r init, echo=FALSE, message=FALSE, warning=FALSE}
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(dpi = 144, fig.width= 6, fig.height = 4, out.width = "90%", fig.align = "center")
```

This vignette demonstrates the functionality of ppstat. It uses
the simulated toyData to illustrate how some toy models can be 
specified and how the fitted models can be visualized. 

```{r, fig.height = 6}
library(ppstat)
data(toyData)
plot(toyData)
```

Standard example of fitting a linear filter model with a filter
function with support $[0,2]$, a linear parametrization and a
basis expansion in terms of B-spline basis functions with knots
equidistantly distributed from $-1$ to $2$.

```{r }
toyPPM <- pointProcessModel(BETA ~ bSpline(ALPHA, knots = seq(-1, 2.5, 0.5)),
                            data = toyData,
                            family = Hawkes(),
                            support = 2)
summary(toyPPM)
termPlot(toyPPM)
```

Alternative analysis. The filter function is composed of two
constant components.

```{r }
toyPPM <- pointProcessModel(BETA ~ cut(ALPHA, c(0, 1, 2),
                                       include.lowest = TRUE),
                            data = toyData,
                            family = Hawkes(),
                            support = 2)
summary(toyPPM)
termPlot(toyPPM)
```

Use of `I()` is required to make constructs like `(ALPHA <= 1)` work
correctly!

```{r }
toyPPM <- pointProcessModel(BETA ~ I(ALPHA <= 1),
                            data = toyData,
                            family = Hawkes(),
                            support = 2)
summary(toyPPM)
termPlot(toyPPM)
```

But the behavior above is perhaps not as expected. The result is a
model equivalent to the model using `cut()` above. To get
the indicator function for the interval $[0,1]$, use the formula
below.

```{r }
toyPPM <- pointProcessModel(BETA ~ as.numeric(ALPHA <= 1),
                            family = Hawkes(),
                            data = toyData,
                            support = 2)
summary(toyPPM)
termPlot(toyPPM)
```

We can include self-dependence of the BETA-points, and a time trend.

```{r }
toyPPM <- pointProcessModel(BETA ~ time + bSpline(ALPHA, knots = seq(-1, 2.5, 0.5)) +
                            bSpline(BETA, knots = seq(-1, 2.5, 0.5)),
                            data = toyData,
                            family = Hawkes(),
                            support = 2)
summary(toyPPM)
termPlot(toyPPM)
```

We can change the `link()` function to the log-link.

```{r }
toyPPM <- pointProcessModel(BETA ~ time + bSpline(ALPHA, knots = seq(-1, 2.5, 0.5)) +
                            bSpline(BETA, knots = seq(-1, 2.5, 0.5)),
                            data = toyData,
                            family = Hawkes("log"),
                            support = 2)                       
summary(toyPPM)
termPlot(toyPPM, trans = exp)
```

The Gibbs family allows for anticipation in the linear filter.

```{r }
toyPPM <- pointProcessModel(BETA ~ bSpline(ALPHA, knots = seq(-2.5, 2.5, 0.5)),
                            data = toyData,
                            family = Gibbs(),
                            support = c(-2, 2))
summary(toyPPM)
termPlot(toyPPM)
```

For the Gibbs family the filter function can be made symmetric, if the
basis functions are. Using `bSpline()` this is achieved as follows.

```{r }
toyPPM <- pointProcessModel(BETA ~ bSpline(ALPHA, knots = seq(-2.5, 2.5, 0.5), sym = TRUE),
                            data = toyData,
                            family = Gibbs(),
                            support = c(-2, 2))
summary(toyPPM)
termPlot(toyPPM)
```

For the model to be a valid model of the point process,
self-dependence terms are not allowed to be anticipating. It is
valid for other terms to be anticipating, and using the Gibbs
family we can specify such a model. 

WARNING: When using the Gibbs
family it is entirely up to the user to make sure that the filter
function for the self-dependence term is 0 on the negative
axis. Using `bSpline()` we set `trunc = 0` to truncate the basis
expansion at 0.

```{r }
toyPPM <- pointProcessModel(BETA ~ bSpline(BETA, knots = seq(-2.5, 2.5, 1), trunc = 0) +
                            bSpline(ALPHA, knots = seq(-2.5, 2.5, 1), trunc = c(-2, 2)),
                            data = toyData,
                            family = Gibbs(),
                            support = c(-2, 2))
summary(toyPPM)
termPlot(toyPPM)
```

Multivariate models can by specified by letting the left
hand side of the formula be a vector of symbols. There is 
a stepwise model selection algorithm for selecting a submodel
based on an information criteria.

```{r }
toyPPM <- pointProcessModel(c(ALPHA, BETA) ~ 
                              bSpline(ALPHA, knots = seq(-0.5, 2.5, 0.5)) +
                              bSpline(BETA, knots = seq(-0.5, 2.5, 0.5)),
                            data = toyData,
                            family = Hawkes("logaffine"),
                            support = 2)
termPlot(toyPPM)
toyPPM <- stepInformation(toyPPM)
termPlot(toyPPM)
```

Terms can also be specified using a smoother and a penalty parameter.

```{r }
toyPPM <- ppSmooth(BETA ~ s(ALPHA) + 
                     bSpline(BETA, knots = seq(-0.5, 2.5, 0.5)),
                   data = toyData,
                   family = Hawkes("logaffine"),
                   support = 2,
                   lambda = 0.1)
termPlot(toyPPM)

toyPPM <- ppKernel(BETA ~ k(ALPHA) + 
                     bSpline(BETA, knots = seq(-0.5, 2.5, 0.5)),
                   data = toyData,
                   family = Hawkes("logaffine"),
                   support = 2,
                   lambda = 0.01)
termPlot(toyPPM)
```

